\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}
I would like to express my sincere gratitude to my parents \textbf{Catia} and \textbf{Fabrizio}, and brother \textbf{Stefano} for their love and support. Your constant encouragement and belief in my abilities have been my greatest source of inspiration. You saw all the struggles and all the victories on my path, and you have stood by my side, never stopping to cheer for me.

I am also grateful to my loving boyfriend, \textbf{Alessandro}, for his unconditional support and presence during the highs and lows of my academic pursuits. You have been my rock during the most challenging times of my studies. You have never doubted my skills even when I was the first one not believing in myself. Having you by my side all along this journey was my biggest treasure.

Furthermore, I cannot express enough gratitude to my exceptional colleagues at the university. Together, we spent countless hours working on assignments and projects, and your genuine support and encouragement have made all the difference. A special thank goes to \textbf{Angelo}, \textbf{Luca}, \textbf{Schiaro}, \textbf{Andrea} and again, \textbf{Alessandro}.

To my dear friends \textbf{Francesca}, \textbf{Sofia}, and \textbf{Dariana}, thank you for making the journey enjoyable with your endless hours of companionship and laughter. Your friendship has been a bright light that lifted the weight of my studies.

My esteemed thesis supervisor, the professor \textbf{Danilo Pianini}, deserves a special mention for his invaluable guidance, unwavering support, and expertise throughout my research journey. Your mentorship and advices have been crucial to the successful completion of this project.

Finally, I extend gratitude to myself for never giving up, even when it felt impossible. This achievement proves my determination and resilience, and I am proud of myself for persevering through the toughest times. This is the end of an incredible journey, and hopefully the beginning of a new beautiful one, which I begin with the certainty that I can achieve incredible goals, such as this one.